Thread Pool in GeckoDB/BOLSTER - Implementing a Thread Pool for a High Performance Database System \\

### 2018-07-03 Presentations/Paper/paper_thread_pool.tex
Small paper fixes

\begin{abstract}
        Processing tasks in parallel is used in nearly all applications to keep
        up with the requirements of modern software systems. However, the
        current implementation of parallel processing in GeckoDB requires
        spawning many short-lived threads that execute a single task and then
        terminate. This creates a lot of overhead since the threads are not
        being reused. To counter this effect we implemented a thread pool to be
        used in combination with GeckoDB to reduce the additional time required
        to create and close thousands of threads. In this paper, we show our
        implementation of a thread pool to process independent tasks in parallel
        including the waiting for a set of processed tasks. We also compare the
        thread pool implementation against the current one. Additionally, we show that the task and thread pool configuration
        depending on the use case has a high impact on the thread
        pool performance. At the end of our evaluation, we show that the implementation 
        fulfils all given constraints and generally reduce
        overhead compared to creating and spawning threads individually.
\end{abstract}

@@

\section{Implementation}
In this section, we describe in detail the design and implementation of our thread pool. We show how tasks are enqueued regarding their priority and how the waiting for tasks is implemented. Additionally, we state our thread metrics and show how they are integrated in the thread pool design.

@@

In Figure 1, we show our design of the whole thread pool system, containing the thread pool itself, the task priority queue and the performance monitoring. Since measuring performance metrics lead to memory and CPU overhead, we decided to exclude the performance measurements from the thread pool to make it optional. The performance monitoring can be activated as a boolean parameter to the thread pool create functions. Consequently, in the target database system, the designers can decide for each thread pool instance, if performance monitoring should be applied.

@@

\subsection{Processing Tasks}
Processing tasks in a thread pool is a balancing act between generalization the of tasks and the performance of their execution. With higher generalization, the implementation can be used for more use cases, but the execution performance may decrease. Therefore, we provide a set of functions to process task in the thread pool to match different requirements of BOLSTER. In Table 1, we show the functions provided by our thread pool to process tasks. One of the requirements of BOLSTER is waiting until a set of tasks is finished. To achieve the same performance as the baseline implementation reaches, we let the main thread first participate in the execution of the tasks and then wait until all tasks with the same handle are finished.

@@

In the next step, the tasks will be enqueued into the priority queue of the thread pool. Before enqueueing, the function \emph{thread\_pool\_enqueue\_tasks\_wait} creates a handle and links all tasks to it. After enqueueing the tasks to the priority queue, the waiting threads in the thread pool pop the enqueued tasks from the priority queue and execute the routine of the task. This happens asynchronously to the calling function and therefore is not considered as an own step of the process. The calling thread gets one of the passed tasks and processes it. After finishing the task, the calling thread waits until all other tasks referring to the created handle are finished and then returns to the calling function. In the next section, we show the waiting for tasks with handles in detail.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section 2, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates. \\
 This approach enables efficient dynamic and lazy resizing of the thread pool. When the number of threads is reduced, the now redundant threads are marked to finish after completing their current task. If later on the number is increased again, still running but marked threads can be enabled again ensuring that few threads have to be created and that the number of running threads will not temporarily exceed the requested size. \\
 After presenting the enqueueing, processing, and waiting for tasks, we show the second component of our thread pool system, the performance monitoring.

@@

\subsection{Testing and Memory Check}
As last aspect of our implementation, we show how we test the thread pool and how we ensure to avoid memory leaks. Since geckoDB already uses GTest \cite{sen2010quick} for unit testing, we also wrote our test using this framework. GTest provides a simple, powerful library to perform unit tests in C++. We wrote unit tests for the priority queue, the thread pool itself and the performance monitoring. As mentioned before, we use functions from the stdatomic library of c. Using this library in GTest results in compiler errors with GCC, since the c atomic implementation differs from the c++ one of the standard library and the GCC developers will not fix this issue. Consequently, we wrote a workaround not using the c stdatomic library in the GTest compilation unit.

@@

\subsection{Evaluation Setup}
In our evaluation setup, we focus primarily on the comparison of using our thread pool against the baseline implementation of creating new threads at runtime of BOLSTER. We simulate calls of BOLSTER primitives with different thread pool and task configurations and compare the results against the baseline.

Additionally, we evaluate the metrics for threads, tasks and the thread pool mentioned in Section 3.F regarding the thread pool and task configuration used in the BOLSTER primitives. To evaluate optimal configurations for the thread pool, we consider the relation between idle and busy time of the thread pool and the average waiting time of tasks.

@@

In our setups mostly Sky- and Kabylake processors have been used with memory
ranging from 8 to 24 GB of both DDR3-SODIMM and DDR4-SODIMM.

All benchmarks have been repeated three times and a median value of all
repetitions is being used for evaluation.

@@

\subsubsection{Benchmark Implementation}
Since BOLSTER is deeply integrated into geckoDB and the workload and usage of multithreading depends on the graph data in the system, we can not directly use the BOLSTER implementation for our evaluation. Thus, we extracted the respecting code parts into an own implementation and use this for a direct comparison against the thread pool.

In Figure \ref{fig4}, we show the approximated version of the BOLSTER multithreading code that we use as baseline implementation. In our benchmark. the method \emph{cmp\_baseline} is called multiple times with a different \emph{numThreads} parameter. Within each call, \emph{numThreads} threads are created and all of these execute the benchmark method v, that calculates a huge amount of exponential values in a loop. After the calling thread takes part in execution, it waits until all created threads are joined. The elapsed processing time is measured from the begin of the method until joining all threads finished.

@@

			clock_gettime(CLOCK_MONOTONIC, &end);
			test_results_baseline[numThreads] +=
				__get_time_diff(&begin, &end);
		}
	\end{lstlisting}
	\caption{Baseline multithreading implementation extracted from BOLSTER}
	\label{fig4}
\end{figure}

@@

In Figure \ref{fig6}, we show the comparison results of both mentioned implementations and the different \emph{numThreads} parameters. Our thread pool implementation is faster than the baseline for all \emph{numThreads} parameters. Based on Figure \ref{fig6}, we make the following conclusions.

Firstly, compared to strict processing workload, create, and later on join threads, consumes much time. Our thread pool implementation reaches up to a multiple of performance increase compared to the baseline. Since the creation and join of threads is the main difference between both tests, reusing threads will lead to this noticeable performance increase.

Secondly, the task creation and enqueuing have only slight on the performance, since the baseline does not include these operations, but is still clearly faster than the thread pool implementation.

@@

\subsection{Summary}
As last step of the evaluation, we summarize our results and compare them to the requirements of geckoDB. As mentioned before, our implementation reaches better performance than the baseline regarding our specific tests. Consequently, we solved the task for implementing and integrating the thread pool into BOLSTER to reach at least the baseline performance. Additionally, using our thread pool implementation solves the problem, that debugging and profiling with a huge number of threads is not possible.

Although a small and constant waiting time was not a requirement for integration into geckoDB, we ensure that assigning many tasks work fast for different thread pool sizes. Regarding the performance and also the design requirements of geckoDB and BOLSTER, we fulfil these requirements and successfully implemented and evaluated our thread pool implementation.

@@

\section{Conclusion and Future Work}
In this work, we describe our implementation of a thread pool system, present in detail the implementation and show our evaluation results containing a comparison against the baseline implementation and the waiting time behaviour over different thread pool sizes. Within our implementation, we state the differences of parallel memory access as well as the problem of implementing synchronous waiting for a set of tasks.

@@

To sum up, our thread pool implementation fulfils all performance and design requirements of geckoDB and BOLSTER and reaches even faster execution time than the existing implementation. In the future, we plan to make the following adaptations to our thread pool system. Firstly, we want to reduce the number of dynamic allocation, since heap allocations mostly are slower than stack allocations. Secondly, we want to examine the usage of existing priority queue implementations for our thread pool system. Thirdly, we plan to integrate a dynamic resizing function based on the collected statistics of the performance monitoring component.

>>>>

\begin{abstract}
        Processing tasks in parallel is used in nearly all applications to keep
        up with the requirements of modern software systems. However, the
        current implementation of parallel processing in GeckoDB requires
        spawning many short-lived threads that execute a single task and then
        terminate. This creates a lot of overhead since the threads are not
        being reused. To counter this effect we implemented a thread pool to be
        used in GeckoDB to reduce the additional time required
        to create and close thousands of threads. In this paper, we show our
        implementation of a thread pool to process independent tasks in parallel,
        including the waiting for a set of processed tasks. We also compare the
        thread pool implementation against the current one. Additionally, we show that the task and thread pool configuration,
        depending on the use case, has a high impact on the thread
        pool performance. At the end of our evaluation, we show that the implementation 
        fulfils all given requirements and generally reduce
        overhead compared to creating and spawning threads individually.
\end{abstract}

@@

\section{Implementation}
In this section, we describe in detail the design and implementation of our thread pool system. We show how tasks are enqueued regarding their priority and how the waiting for tasks is implemented. Additionally, we state our thread metrics and show how they are integrated in the thread pool design.

@@

In Figure 1, we show our design of the whole thread pool system, containing the thread pool itself, the task priority queue and the performance monitoring. Since measuring performance metrics lead to memory and CPU overhead, we decided to exclude the performance measurements from the thread pool to make it optional. The performance monitoring can be activated by a boolean parameter of the thread pool create functions. Consequently, in the target database system, the designers can decide for each thread pool instance, if performance monitoring should be applied.

@@

\subsection{Processing Tasks}
Processing tasks in a thread pool is a balancing act between generalization the of tasks and the performance of their execution. With higher generalization, the implementation can be used for more use cases, but the execution performance may decrease. Therefore, we provide a set of functions to process task in the thread pool to match different requirements of BOLSTER. In Table 1, we show the functions provided by our thread pool to process tasks. One of the requirements of BOLSTER is waiting until a set of tasks is finished. To achieve at least the same performance as the baseline implementation reaches, we let the main thread first participate in the execution of the tasks and then wait until all tasks with the same handle are finished.

@@

In the next step, the tasks will be enqueued into the priority queue of the thread pool. Before enqueueing, the function \emph{thread\_pool\_enqueue\_tasks\_wait} creates a handle and links all tasks to it. After enqueueing the tasks to the priority queue, the waiting threads in the thread pool pop the enqueued tasks from the priority queue and execute the task routine. We decided to let threads look for new tasks instead of explicitly assigning to reduce the scheduling overhead. We call this approach \emph{first come first serve}. This happens asynchronously to the calling function and therefore is not considered as an own step of the process. The calling thread gets one of the passed tasks and processes it. After finishing the task, the calling thread waits until all other tasks referring to the created handle are finished and then returns to the calling function. In the next section, we show the waiting for tasks with handles in detail.

@@

\subsection{Managing Thread States}
As mentioned before, accessing the same memory location from different threads may lead to race conditions, which means the program behaviour depends on the accessing order of the threads. For example, while one thread updates data at a specific memory location, another thread writes into the data, which results to inconsistency. In Section 2, we presented two approaches to handle multithreaded access to the same memory location, mutex threads and atomic operations and we state where those approaches are used in our implementation. In this subsection, we focus on atomic operations to manage thread states. We consider the following steps for a thread to update its state. At first, the current state has to be checked and, depending on the current state, it has to be updated. If these steps are implemented using an if statement and an update function, the complete procedure is not atomic. Consequently, another thread can update the state between check of the current state and update. To avoid this, we use the function \emph{atomic\_compare\_exchange\_strong} \cite{atomicstrong} of the stdatomic library. This function combines the check and update in one atomic procedure. We implemented every thread state access with this function to ensure atomic thread state updates.

This approach enables efficient dynamic and lazy resizing of the thread pool. When the number of threads is reduced, the now redundant threads are marked to finish after completing their current task. If later on the number is increased again, still running but marked threads can be enabled again ensuring that few threads have to be created and that the number of running threads will not temporarily exceed the requested size. 
 After presenting the enqueueing, processing, and waiting for tasks, we show the second component of our thread pool system, the performance monitoring.

@@

\subsection{Testing and Memory Check}
As last aspect of our implementation, we show how we test the thread pool and how we ensure to avoid memory leaks. Since GeckoDB already uses GTest \cite{sen2010quick} for unit testing, we also wrote our test using this framework. GTest provides a simple, powerful library to perform unit tests in C++. We wrote unit tests for the priority queue, the thread pool itself and the performance monitoring. As mentioned before, we use functions from the stdatomic library of c. Using this library in GTest results in compiler errors with GCC, since the c atomic implementation differs from the c++ one of the standard library and the GCC developers will not fix this issue. Consequently, we wrote a workaround not using the c stdatomic library in the GTest compilation unit.

@@

\subsection{Evaluation Setup}
In our evaluation setup, we focus primarily on the comparison of our thread pool against the baseline implementation in BOLSTER, that creates new threads at runtime for each request. We simulate calls of BOLSTER primitives with different thread pool and task configurations and compare the results against the baseline.

Additionally, we evaluate the metrics for threads, tasks and the thread pool mentioned in Section 3.F regarding the thread pool and task configuration used in the BOLSTER primitives. To evaluate optimal configurations for the thread pool, we consider the relation between waiting and completion time of  thread pool instances with different sizes.

@@

In our setups mostly Sky- and Kabylake processors have been used with memory
ranging from 8 to 24 GB of both DDR3-SODIMM and DDR4-SODIMM. 
All benchmarks have been repeated three times and a median value of all
repetitions is being used for evaluation.

@@

\subsubsection{Benchmark Implementation}
Since BOLSTER is deeply integrated into GeckoDB and the workload and usage of multi threading depends on the graph data in the system, we can not directly use the BOLSTER implementation for our evaluation. Thus, we extracted the respecting code parts into an own implementation and use this for a direct comparison against one that uses our thread pool.

In Figure \ref{fig4}, we show the approximated version of the BOLSTER multi threading code that we use as baseline implementation. In our benchmark. the method \emph{cmp\_baseline} is called multiple times with a different \emph{numThreads} parameter. Within each call, \emph{numThreads} threads are created and all of these execute the benchmark method \emph{work\_large}, that calculates a huge amount of exponential values in a loop. After the calling thread takes part in execution, it waits until all created threads are joined. The elapsed processing time is measured from the begin of the method until the join of all threads finished.

@@

			clock_gettime(CLOCK_MONOTONIC, &end);
			test_results_baseline[numThreads] +=
				__get_time_diff(&begin, &end);
		}
	\end{lstlisting}
	\caption{Baseline multi threading implementation extracted from BOLSTER}
	\label{fig4}
\end{figure}

@@

In Figure \ref{fig6}, we show the comparison results of both mentioned implementations and the different \emph{numThreads} parameters. Our thread pool implementation is faster than the baseline for all \emph{numThreads} parameters. Based on Figure \ref{fig6}, we make the following conclusions.  
Firstly, compared to strict processing workload, create, and later on join threads, consumes much time. Our thread pool implementation reaches up to a multiple of performance increase compared to the baseline. Since the creation and join of threads is the main difference between both tests, reusing threads will lead to this noticeable performance increase.

Secondly, the task creation and enqueuing have only slight on the performance, since the baseline does not include these operations, but is still clearly slower than the thread pool implementation.

@@

\subsection{Summary}
As last step of the evaluation, we summarize our results and compare them to the requirements of GeckoDB. As mentioned before, our implementation reaches better performance than the baseline regarding our specific tests. Consequently, we solved the task for implementing and integrating the thread pool into BOLSTER to reach at least the baseline performance. Additionally, using our thread pool implementation solves the problem, that debugging and profiling with a huge number of threads is not possible.

Although a small and constant waiting time was not a requirement for integration into GeckoDB, we ensure that assigning many tasks works fast for different thread pool sizes. Regarding the performance and also the design requirements of GeckoDB and BOLSTER, we fulfil these requirements and successfully implemented and evaluated our thread pool implementation.

@@

\section{Conclusion and Future Work}
In this work, we describe our implementation of a thread pool system. We present in detail the implementation and show our evaluation results containing a comparison against the baseline implementation and the waiting time behaviour over different thread pool sizes. Within our implementation, we state the differences of parallel memory access as well as the problem of implementing synchronous waiting for a set of tasks.

@@

To sum up, our thread pool implementation fulfils all performance and design requirements of GeckoDB and BOLSTER and reaches even faster execution time than the existing implementation. In the future, we plan to make the following adaptations to our thread pool system. Firstly, we want to reduce the number of dynamic allocation, since heap allocations mostly are slower than stack allocations. Secondly, we want to examine the usage of existing priority queue implementations for our thread pool system. Thirdly, we plan to integrate a dynamic resizing function based on the collected statistics of the performance monitoring component.
